Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-03-11 paper/paper1/ms.tex
More surgery and editting to the methods section, much more coherent

The computational bottleneck occurs at the evaluation of Equation \ref{equation1}, which can be viewed as having a $N_{\mathrm{lines}}\times N_{S}$ matrix $\bm{\bar{V}}$ assembled by stacking each Voigt absoption profile $\mathsf{V}_j(\bm{\lambda}_s)$ on top of each other:

@@

Equation \ref{equation1} performs a type of matrix contraction, turning a $N_{\mathrm{lines}}\times N_{S}$ matrix into a $1\times N_{S}$ vector. The number of Floating Point Operations (FLOPS) scales with the number of entries in this matrix. Efficient GPU algorithms exist for matrix manipulations such as this one, allowing this computation to proceed quickly on modern machines. In particular, the proprietary CUDA architecture for NVIDIA\textsuperscript{\tiny\textregistered} GPUs contains Tensor cores with specialized matrix math. The chief bottleneck occurs when the storage of the $\bm{\bar{V}}$ matrix exceeds the available RAM of a GPU or CPU: the computation will fail with an ``Out of Memory'' exception. Modern NVIDIA GPUs have $8-40$ GB of RAM. The memory bottleneck is even more pernicious than mere storage of the $\bm{\bar{V}}$ matrix since the CPU/GPU also has to store the computation graph for autodiff. It is generally not possible to evaluate Equation \ref{equation1} in its entirety in one-fell-swoop. A remedy is needed.

@@

Another solution to the computation problem is to take advantage of the sparsity of the $\bm{\bar{V}}$ matrix: most of the Voigt profile's entries far from the line center are vanishingly close to zero. Ideally, we would only evaluate the model close to the line center position, some number of line cuts $N_{\mathrm{cut}}=\alpha \cdot \mathrm{FWHM}$ away. These line cuts produce a speedup by a factor of $\frac{N_s}{N_{\mathrm{cut}}}$, which can exceed $100\times$ for wide bandwidth spectra. Furthermore, efficient algorithms for assembling and coalescing sparse matrices exist in \texttt{PyTorch}. A few practical considerations make sparse matrices peskier to work with than their dense counterparts. First, the $\alpha$ hyperparameter should be set large enough that the truncation effect is not seen for the deepest lines, or in post-processing steps such as line broadening. The $\alpha$ should be preselected before the training process. If the FWHM initialization was poor enough, the question may arise how autodiff training will cope with the poor performance of a model, leading to the prospect of strange and difficult-to-diagnose artifacts near spurious line-cut truncation artifacts.

We implemented sparse matrices in \emph{blas\'e} using the \texttt{sparse\_coo\_tensor} function in \emph{PyTorch}.

There are a few standardization strategies for applying wing cuts. Here we coerce all wing cuts to be the same number of pixels, typically 6000 pixels for \texttt{PHOENIX}, with the middle pixel being at the line center position, and about 3000 pixels to the red and blue side of the line. Most spectral lines only appreciably affect the signal in the central $\sim5-50$ pixels, so such a large wing cut may seem over-zealous. We emphasize that the line cut serves as the fixed pixel locations where the line will be evaluated with post-processing steps such as radial velocity shifts and line broadening. So these 6000 pixels must be capable of handling the fastest-moving and fastest-spinning stars you expect to see in your observed spectra of interest.

The PHOENIX spectra double their pixel sampling shortward of $\lambda = 1 \mu$m, so 6000 pixels in the visible is about 30 Angstroms, but 6000 pixels in the infrared is about 60 \AA. We \emph{could have} demanded that the wing cut be exactly 30 \AA, or some other heuristic, but that choice would result in so-called \emph{jagged} or \emph{ragged} arrays: collections of 1D arrays that differ in their lengths. Some lines would have 5999 pixels, some would have 6000, and others would have 2999 or 3000 pixels, depending whether they occur to the red or blue side of the $1 \;\mu$m kink in sampling. Most efficient CPU and GPU algorithms are not equipped to deal with ragged arrays, therefore increasing the computation time. Instead, the standardization choice of exactly 6000 pixels means we can use standard matrix or tensor arithmetic, which somewhat counterintuitively is faster while handling more pixels for all spectral lines: 6000 pixels everywhere is faster than 6000 pixels for some and 2999 for other lines.

Locating the indices of these 6000 ``nearest-neighbor'' pixels is very fast, only occurs once, and then is fixed.

We rewrite Equation \ref{equation1} as a sum by taking the log of both sides

\begin{eqnarray}
    \ln{\mathsf{F}(\bm{\lambda})} = \sum_{i=1}^{N_{lines}} \ln{\left(1-V(\bm{\lambda};A_i, \lambda_{c,i}, \sigma_i) \right)}
\end{eqnarray}

The argument of the natural logarithm approaches 1 far from the line core, making the terms outside of the matrix approximately zero, a sparse matrix representing the log of the flux. Sparse matrix methods generally support an operation known as \emph{coalescing}, which sums values with repeated indices. The coalescing operation is extremely fast and efficient on either CPU or GPU hardware. We sum the logs and then exponentiate the result before comparing it to the native linear flux. We enforce that the argument of the natural log must be greater than zero through a simple truncation, which may approximately resemble the effect of saturation in line cores. This truncation would only be relevant for saturated lines. Each pixel may get computed about $\sim100$ times in this sparse implementation, which is about $50\times$ better than each pixel getting computed $N_{lines}\sim6500$ times in the dense approach.

Some lines, such as Hydrogen, Sodium, potassium, neutral metal lines, and others have extremely broad line wings, approaching larger than the $\sim6000$ pixels we allocate for the sparse implementation. These special lines need to be handled separately from the weak lines, both from a computational performance perspective and an accuracy perspective.

@@

>>>>

The computational bottleneck occurs at the evaluation of Equation \ref{equation1}, which can be viewed as having a $N_{\mathrm{lines}}\times N_{S}$ matrix $\bm{\bar{F}}$ assembled by stacking each Voigt absoption profile $\mathsf{V}_j(\bm{\lambda}_s)$ on top of each other:

@@

An element of this matrix, $F_{j\lambda}$, will have the flux value for a given line at a given wavelength coordinate. Equation \ref{equation1} performs a type of matrix contraction, turning a $N_{\mathrm{lines}}\times N_{S}$ matrix into a length $N_{S}$ row vector. The number of Floating Point Operations (FLOPS) scales with the number of entries in this matrix. We can rewrite Equation \ref{equation1} as a sum by taking the log of both sides and dropping in this $\bm{\bar{F}}$ matrix:

\begin{eqnarray}
    \ln{\mathsf{S}_{\rm clone}} = \sum_{j=1}^{N_{lines}} \ln{F_{j\lambda}} = \mathbf{1} \cdot \ln{\bm{\bar{F}}}
\end{eqnarray}

where $\mathbf{1}$ is a $1\times N_{\rm lines}$ row vector. Efficient GPU algorithms exist for matrix manipulations such as this one, allowing this computation to proceed quickly on modern machines. In particular, the proprietary CUDA architecture for NVIDIA\textsuperscript{\tiny\textregistered} GPUs contains Tensor cores with specialized matrix math. The chief bottleneck occurs when the storage of the $\bm{\bar{V}}$ matrix exceeds the available RAM of a GPU or CPU: the computation will fail with an ``Out of Memory'' exception. Modern NVIDIA GPUs have $8-40$ GB of RAM. The memory bottleneck is even more pernicious than mere storage of the $\bm{\bar{V}}$ matrix since the CPU/GPU also has to store the computation graph for autodiff. It is generally not possible to evaluate Equation \ref{equation1} in its entirety in one-fell-swoop. A remedy is needed.

@@

The $\ln{\bm{\bar{F}}}$ matrix is sparsely populated: most of the entries far from the line center are vanishingly close to zero. Here we take advantage of that mostly empty matrix using the mathematics of sparse matrices \citep{saad03:IMS}.

We retain a relatively small number of pixels $N_{\rm cut}$ adjacent to the line center. Setting this wing cut produces a speedup by a factor of $\frac{N_S}{N_{\mathrm{cut}}}$, which can exceed $100\times$ for wide bandwidth spectra.The choice of $N_{\rm cut}$ is nuanced.  It should be set large enough that truncation effects are not seen for the broadest lines.  But even more, $N_{\rm cut}$ has to be future-proofed for Doppler-shifting. Extreme Doppler shifts could hypothetically send line cores entirely outside the extents of $N_{\rm cut}$.  We therefore typically set wingcuts comparable to the buffer size $\pm \Delta \lambda_{\mathrm{buffer}}$, even though most weak lines only perceptibly affect $<1\;$\AA. We coerce all wing cuts to be the same number of pixels, typically 6000 pixels for \texttt{PHOENIX}, with the middle pixel being at the line center position, and about 3000 pixels to the red and blue side of the line. We refer to this approximate sparse matrix as $\ln{\bm{\hat{F}}}$.

The PHOENIX spectra double their pixel sampling shortward of $\lambda = 1 \mu$m, so 6000 pixels in the visible is about 30 Angstroms, but 6000 pixels in the infrared is about 60 \AA. We \emph{could have} demanded that the wing cut be exactly 30 \AA, or some other heuristic, but that choice would result in so-called \emph{jagged} or \emph{ragged} arrays: collections of 1D arrays that differ in their lengths. Some lines would have 5999 pixels, some would have 6000, and others would have 2999 or 3000 pixels, depending whether they occur to the red or blue side of the $1 \;\mu$m kink in sampling. Most efficient CPU and GPU algorithms are not equipped to deal with ragged arrays, therefore increasing the computation time. Instead, the standardization choice of exactly 6000 pixels means we can use standard matrix or tensor arithmetic, which somewhat counterintuitively is faster while handling more pixels for all spectral lines: 6000 pixels everywhere is faster than 6000 pixels for some and 2999 for other lines.  Locating the indices of these 6000 ``nearest-neighbor'' pixels is very fast, only occurs once, and then is fixed.

Sparse matrix methods generally support an operation known as \emph{coalescing}, which sums values with repeated indices.  We sum the logs and then exponentiate the result before comparing it to the native linear flux. We enforce that the argument of the natural log must be greater than zero through a simple truncation, which may approximately resemble the effect of saturation in line cores. This truncation would only be relevant for saturated lines. Each pixel may get computed about $\sim100$ times in this sparse implementation, which is about $50\times$ better than each pixel getting computed $N_{lines}\sim6500$ times in the dense approach.  Efficient algorithms for assembling and coalescing sparse matrices exist in \texttt{PyTorch}

@@

Some lines, such as Hydrogen, Sodium, potassium, neutral metal lines, and others have extremely broad line wings, approaching larger than the $\sim6000$ pixels we allocate for the sparse implementation. These special lines should be handled separately from the weak lines, both from a computational performance perspective and an accuracy perspective.
